\huge{Meta-analytic evidence that sexual selection improves population fitness}

### 2019-04-05 Response_to_editorial_requests.Rmd

\begin{itemize}

\item The title is now \textit{Meta-analytic evidence that sexual selection improves population fitness}.

\item In the abstract we include: \textbf{We find evidence} that sexual selection on males tends to elevate the mean and reduce the variance for many fitness traits.

\item In the abstract we include: However, sexual selection had weaker positive effects on direct measures of population fitness, such as extinction rate and proportion of viable offspring, than on individual fitness measures with indirect or less clear relationships to population fitness.

\item We have modified the 1st paragraph of the discussion and make it clear which conclusions are more or less strongly supported by the data (see also following comments for additional changes to improve clarity of interpretation).

### 2019-04-05 Manuscripts/Manuscript_MetaAnalysis.Rmd
Minor tweaks

---
title: '\huge{Meta-analytic evidence that sexual selection improves population fitness}'
author: 
- Justin G. Cally\textsuperscript{§*}, Devi Stuart-Fox\textsuperscript{§} and Luke Holman\textsuperscript{§}
- '\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia'
- '\textsuperscript{*}\textit{justin.g.cally@gmail.com}'
output:
  pdf_document: default
  html_document: default
  word_document:
      pandoc_args: ["--smart"]
bibliography: Bibliography/Bibliography.bib
subparagraph: yes
urlcolor: blue
csl: Bibliography/NComms.csl
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection improves fitness} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers \usepackage{setspace}\onehalfspacing
    \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{} \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{} \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} 
    \usepackage{hyperref}\hypersetup{colorlinks=true, linkcolor=blue, filecolor=blue, urlcolor=blue} 
    \usepackage{color,soul}
    \raggedbottom    
abstract: Sexual selection has manifold ecological and evolutionary consequences, making its net effect on population fitness difficult to predict from first principles. Instead, one can empirically determine whether sexual selection increases or decreases population fitness by manipulating the opportunity for sexual selection in replicate populations, allow evolution to proceed, and then compare the evolved average fitness between the treatment groups. Here, we synthesise 459 effect sizes from 65 pertinent experimental evolution studies using meta-analysis. We find evidence that sexual selection on males tends to elevate the mean and reduce the variance for many fitness traits, especially in females evolving under stressful conditions. However, sexual selection had comparatively less positive effects on traits directly related to fitness than those with indirect or ambiguous relationships. Ultimately, we conclude that the beneficial population-level consequences of sexual selection typically outweigh the harmful ones and that the effects of sexual selection on total selection, genetic variance, and/or evolvability can differ markedly between sexes and environments. We discuss the implications of these results for conservation and evolutionary biology.
---
\pagestyle{plain}

@@

The grand mean across all types of effect sizes (direct, indirect and ambiguous) was positive (REML $\beta$ = 0.24, 95% CIs: 0.055 to 0.43, $p$ = 0.011; Bayesian $\beta$ = 0.25, 95% CIs: -0.0074 to 0.51, $BF_{>0}$ = 35), indicating that sexual selection on males typically had a net positive effect on the majority of populations and fitness components so far studied. Moreover, the effect sizes associated with the manipulation of sexual selection varied between different fitness traits. Sexual selection had a beneficial effect on most fitness traits, but varied across the three relationships to fitness (Figure 1; \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_on_direct,_indirect_and_ambiguous_measures_of_fitness}{Supplementary Table 6}). Sexual selection elevated fitness for traits that shared an ambiguous relationship to fitness (REML $\beta$ = 0.21, 95% CIs: 0.058 to 0.093; Bayesian $\beta$ = 0.20, 95% CIs: -0.0016 to	0.39, $BF_{>0}$ = 38, $n$ = 144) and an indirect relationship to fitness (REML $\beta$ = 0.24, 95% CIs: 0.13 to 0.36; Bayesian $\beta$ = 0.24, 95% CIs: 0.033 to 0.43, $BF_{>0}$ = 59, $n$ = 141). Additionally, sexual selection elevated fitness components directly related to fitness, albeit at a lower magnitude (REML $\beta$ = 0.13, 95% CIs: 0.019 to 0.24; Bayesian $\beta$ = 0.13, 95% CIs: -0.079 to 0.31, $BF_{>0}$ = 11, $n$ = 174). A large forest plot with predicted effect sizes for each fitness component is presented in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Figure 1} and further detailed with model predictions and individual meta-analyses in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Table 8,9}. Sexual selection significantly reduced two fitness components, namely immunity (REML $\beta$ = -0.42, 95% CIs: -0.64 to -0.20; Bayesian $\beta$ = -0.43, 95% CIs: -0.70 to -0.15, $BF_{>0}$ = 0.0026; $n$ = 35) and body condition (REML $\beta$ = -1.2, 95% CIs: -1.8 to -0.63; Bayesian $\beta$ = -1.2, 95% CIs: -1.9 to -0.63, $BF_{>0}$ < 0.0001, $n$ = 1).  \phantomsection

@@

We found that the sex of the individuals measured (male, female, or a mixture), and the conditions under which the population evolved (stressful or benign) interacted to affect the relationship between sexual selection and fitness (Table 1,2). Sexual selection on males significantly improved female fitness, and the beneficial effect of sexual selection was significantly stronger for females from populations evolving under stressful conditions (*e.g.* a food source to which they were not well-adapted) than under benign conditions (Figure 2a, Table 2). Sexual selection had a positive but non-significant effect on male fitness, and in contrast to females, fitness benefits were significantly weaker in stressful than benign environments (Figure 2a, Table 2). Consistent with the different consequences of sexual selection for female and male fitness, the mean effect size in mixed-sex samples was non-significantly positive, and there was no significant difference between benign versus stressful conditions (Figure 2a, Table 2). Overall, our results indicate that the positive effect of sexual selection on fitness is greater for females than males, and the difference between the sexes is magnified in stressful environments. When we used an alternative measure of effect size (log response ratio or lnRR), the results aligned with those of Hedges' g in that sexual selection elevates population fitness, with its effect magnified for females evolving in stressful environments (\href{https://justincally.github.io/SexualSelection/#meta-analysis_using_lnrr}{Supplementary Figure 4 and Supplementary Table 16}).

@@

By applying meta-analysis to log coefficient of variation ratios [*lnCVR*; @RN177], we found evidence that sexual selection reduces phenotypic variation under certain conditions (Figure 2b). Specifically, phenotypic variance was significantly reduced by sexual selection for fitness components measured in females under stressful conditions (*lnCVR* = -0.78, 95% CIs: -1.23 to -0.34, $n$ = 27). By contrast, we found no significant effect of sexual selection on phenotypic variance in males, or for either sex under benign conditions (Figure 2b; \href{https://justincally.github.io/SexualSelection/#meta-analysis_on_variance}{Supplementary Tables 17-20}). However, similar to the results in females, there was a non-significant trend for a reduction in phenotypic variance in mixed-sex samples measured under stressful conditions (*lnCVR* = -0.76, 95% CIs: -1.22 to -0.31; Figure 2b). A meta-analysis using the log variability ratio (lnVR), which does not account for the mean-variance relationship present in the dataset (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 5}), suggested sexual selection reduces variance for mixed-sex samples in stressful conditions, but not females, results of this meta-analysis can be found in the Supplementary Information (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 6 and Supplementary Table 21}).

@@

Our meta-analysis revealed that populations evolving under sexual selection often have higher values for multiple fitness traits, relative to populations where sexual selection on males was experimentally removed or weakened. Sexual selection had beneficial effects for traits that were ambiguously or indirectly related to fitness, but had smaller and variable effects on direct measures of fitness such as extinction rate, reproductive success (defined as the number of offspring produced), and the proportion of viable offspring (similar to the results of another meta-analysis [@Prokop_2012]). Fitness traits related to immunocompetence were an exception: sexual selection typically resulted in weaker immunity. This result is interesting in light of the hypothesised trade-off between sexually-selected phenotypes and immunity, for example, due to immunosuppressive effects of sex hormones [@RN323; @RN324]. Furthermore, the overall benefit of sexual selection was greater for females than males, and this sex difference was magnified in stressful environments. Consistent with stronger selection on female fitness under stress, female and mixed-sex samples showed reduced phenotypic variance when sexual selection was applied under stressful as opposed to benign conditions. These results suggest that sexual selection may contribute to population persistence under stressful conditions, such as fluctuating environmental change [@RN173] or spatial variability [@RN189], particularly since female reproductive output is often a limiting factor in population growth [@RN188].

>>>>

---
title: '\huge{Meta-analytic evidence that sexual selection improves population fitness}'
author: 
- Justin G. Cally\textsuperscript{§*}, Devi Stuart-Fox\textsuperscript{§} and Luke Holman\textsuperscript{§}
- '\textsuperscript{§}School of BioSciences, The University of Melbourne, Parkville, VIC, 3052, Australia'
- '\textsuperscript{*}\textit{justin.g.cally@gmail.com}'
output:
  pdf_document: default
  html_document: default
  word_document:
      pandoc_args: ["--smart"]
bibliography: Bibliography/Bibliography.bib
subparagraph: yes
urlcolor: blue
csl: Bibliography/NComms.csl
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual selection improves fitness} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers \usepackage{setspace}\onehalfspacing
    \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{} \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{} \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} 
    \usepackage{hyperref}\hypersetup{colorlinks=true, linkcolor=blue, filecolor=blue, urlcolor=blue} 
    \usepackage{color,soul}
    \raggedbottom    
abstract: Sexual selection has manifold ecological and evolutionary consequences, making its net effect on population fitness difficult to predict from first principles. Instead, one can empirically determine whether sexual selection increases or decreases population fitness by manipulating the opportunity for sexual selection in replicate populations, allow evolution to proceed, and then compare the evolved average fitness between the treatment groups. Here, we synthesise 459 effect sizes from 65 pertinent experimental evolution studies using meta-analysis. We find evidence that sexual selection on males tends to elevate the mean and reduce the variance for many fitness traits, especially in females evolving under stressful conditions. However, sexual selection had weaker positive effects on direct measures of population fitness, such as extinction rate and proportion of viable offspring, than on individual fitness measures with indirect or less clear relationships to population fitness. Overall, we conclude that the beneficial population-level consequences of sexual selection typically outweigh the harmful ones and that the effects of sexual selection on total selection, genetic variance, and/or evolvability can differ markedly between sexes and environments. We discuss the implications of these results for conservation and evolutionary biology.
---
\pagestyle{plain}

@@

The grand mean across all types of effect sizes (direct, indirect and ambiguous) was positive (REML $\beta$ = 0.24, 95% CIs: 0.055 to 0.43, $p$ = 0.011; Bayesian $\beta$ = 0.25, 95% CIs: -0.0074 to 0.51, $BF_{>0}$ = 35), indicating that sexual selection on males typically had a net positive effect on the majority of populations and fitness components so far studied. Moreover, the effect sizes associated with the manipulation of sexual selection varied between different fitness traits. Sexual selection had a beneficial effect on most fitness traits, but varied across the three relationships to fitness (Figure 1; \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_on_direct,_indirect_and_ambiguous_measures_of_fitness}{Supplementary Tables 5,6}). Sexual selection elevated fitness for traits that shared an ambiguous relationship to fitness (REML $\beta$ = 0.21, 95% CIs: 0.058 to 0.093; Bayesian $\beta$ = 0.20, 95% CIs: -0.0016 to	0.39, $BF_{>0}$ = 38, $n$ = 144) and an indirect relationship to fitness (REML $\beta$ = 0.24, 95% CIs: 0.13 to 0.36; Bayesian $\beta$ = 0.24, 95% CIs: 0.033 to 0.43, $BF_{>0}$ = 59, $n$ = 141). Additionally, sexual selection elevated fitness components directly related to fitness, albeit at a lower magnitude (REML $\beta$ = 0.13, 95% CIs: 0.019 to 0.24; Bayesian $\beta$ = 0.13, 95% CIs: -0.079 to 0.31, $BF_{>0}$ = 11, $n$ = 174). A large forest plot with predicted effect sizes for each fitness component is presented in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Figure 1, Supplementary Table 7} and further detailed with model predictions and individual meta-analyses in \href{https://justincally.github.io/SexualSelection/#effect_of_sexual_selection_for_different_fitness_components}{Supplementary Table 8,9}. Sexual selection significantly reduced two fitness components, namely immunity (REML $\beta$ = -0.42, 95% CIs: -0.64 to -0.20; Bayesian $\beta$ = -0.43, 95% CIs: -0.70 to -0.15, $BF_{>0}$ = 0.0026; $n$ = 35) and body condition (REML $\beta$ = -1.2, 95% CIs: -1.8 to -0.63; Bayesian $\beta$ = -1.2, 95% CIs: -1.9 to -0.63, $BF_{>0}$ < 0.0001, $n$ = 1).  \phantomsection

@@

We found that the sex of the individuals measured (male, female, or a mixture), and the conditions under which the population evolved (stressful or benign) interacted to affect the relationship between sexual selection and fitness (Table 1,2 and \href{https://justincally.github.io/SexualSelection/#bayesian_model:_effects_of_environment_and_sex}{Supplementary Tables 11-13}). Sexual selection on males significantly improved female fitness, and the beneficial effect of sexual selection was significantly stronger for females from populations evolving under stressful conditions (*e.g.* a food source to which they were not well-adapted) than under benign conditions (Figure 2a, Table 2, \href{https://justincally.github.io/SexualSelection/#bayesian_model:_effects_of_environment_and_sex}{Supplementary Table 14}). Sexual selection had a positive but non-significant effect on male fitness, and in contrast to females, fitness benefits were significantly weaker in stressful than benign environments (Figure 2a, Table 2). Consistent with the different consequences of sexual selection for female and male fitness, the mean effect size in mixed-sex samples was non-significantly positive, and there was no significant difference between benign versus stressful conditions (Figure 2a, Table 2). Overall, our results indicate that the positive effect of sexual selection on fitness is greater for females than males, and the difference between the sexes is magnified in stressful environments. When only fitness components directly related to fitness were used in the mixed-effects model the benefits of sexual selection were magnified in females evolving under stressful conditions (\href{https://justincally.github.io/SexualSelection/#effects_of_environment_and_sex_on_direct_fitness_components}{Supplementary Figure 3 and Supplementary Table 15}). Similarly, when we used an alternative measure of effect size (log response ratio or lnRR), the results aligned with those of Hedges' g in that sexual selection elevates population fitness, with its effect magnified for females evolving in stressful environments (\href{https://justincally.github.io/SexualSelection/#meta-analysis_using_lnrr}{Supplementary Figure 4 and Supplementary Table 16}).

@@

By applying meta-analysis to log coefficient of variation ratios [*lnCVR*; @RN177], we found evidence that sexual selection reduces phenotypic variation under certain conditions (Figure 2b). Specifically, phenotypic variance was significantly reduced by sexual selection for fitness components measured in females under stressful conditions (*lnCVR* = -0.78, 95% CIs: -1.23 to -0.34, $n$ = 27). By contrast, we found no significant effect of sexual selection on phenotypic variance in males, or for either sex under benign conditions (Figure 2b; \href{https://justincally.github.io/SexualSelection/#meta-analysis_on_variance}{Supplementary Tables 17-20}). However, similar to the results in females, there was a non-significant trend for a reduction in phenotypic variance in mixed-sex samples measured under stressful conditions (*lnCVR* = -0.76, 95% CIs: -1.22 to -0.31; Figure 2b). A meta-analysis using the log variability ratio (lnVR), which does not account for the mean-variance relationship present in the dataset (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 5}), suggested sexual selection reduces variance for mixed-sex samples in stressful conditions, but not females. Results of this meta-analysis can be found in the Supplementary Information (\href{https://justincally.github.io/SexualSelection/#lnvr_effect_size_of_variance}{Supplementary Figure 6 and Supplementary Table 21}).

@@

Our meta-analysis revealed that populations evolving under sexual selection often have higher values for multiple fitness traits, relative to populations where sexual selection on males was experimentally removed or weakened. Sexual selection had beneficial effects on the majority of commonly measured individual fitness traits, which have indirect or unclear relationships to population fitness. Effects of sexual selection on direct measures of population fitness such as extinction rate, reproductive success (defined as the number of offspring produced), and the proportion of viable offspring, were smaller and variable (similar to the results of another meta-analysis [@Prokop_2012]), but nonetheless tended to be positive. Fitness traits related to immunocompetence were an exception: sexual selection typically resulted in weaker immunity. This result is interesting in light of the hypothesised trade-off between sexually-selected phenotypes and immunity, for example, due to immunosuppressive effects of sex hormones [@RN323; @RN324]. Furthermore, the overall benefit of sexual selection was greater for females than males, and this sex difference was magnified in stressful environments. Consistent with stronger selection on female fitness under stress, female and mixed-sex samples showed reduced phenotypic variance when sexual selection was applied under stressful as opposed to benign conditions. These results suggest that sexual selection may contribute to population persistence under stressful conditions, such as fluctuating environmental change [@RN173] or spatial variability [@RN189], particularly since female reproductive output is often a limiting factor in population growth [@RN188].
